Assessing the Global COVID19 Impact on Air Transport with Open Data

### 2021-07-22 paper.Rmd
starting final drafting

---
title: "Assessing the Global COVID19 Impact on Air Transport with Open Data"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
# failed to render and I could not finally track what kills the pdf
#  bookdown::pdf_book:
#    base_format: rticles::ieee_article
  bookdown::word_document2
bibliography: "DASC2021.bib"
affiliation:
  ## use one only of the following
  # author-columnar: true         ## one column per author
  institution-columnar: true  ## one column per institution  
  # wide: true          ## one column wide author/affiliation fields
  institution:
    - name: EUROCONTROL
      department: Performance Review Unit
      location: Brussels (Belgium)
      #mark: 1
      author:
        - name: Rainer Koelle
          email: rainer.koelle@eurocontrol.int
    - name: DECEA
      department: Subdepartment of Operations 
      location: Rio de Janeiro (Brazil)
      #mark: 2
      author:
        - name: Fabio Lourenco Carneiro Barbosa
          email: barbosaflcb@fab.mil.br
abstract: |
  This paper approaches the impact of the pandemic as a massive service disruption of the pre-pandemic global     connectivity and regional air transport networks. In particular, the project aims to provide data analytical evidence for policy success and transformation of the air transportation system. As an aspirational goal, the industry aims to recover in a “greener” manner. The project builds on openly available data sets. The paper will be produced in a reproducible manner making the data, code, and its processing available to interested reseachers and practitioners. The open assessment will provide policy makers with a tool to assess the reaction to local or regional measures.
  <!-- motivation -->
  <!-- short background - why important -->
  <!-- approach chosen / research method -->
  <!-- experimental work -->
  <!-- results obtained -->
  <!-- conclusions / key take-away --> 
---

@@

For several years, many concerns of the global air traffic management community has been directed to the evident problem of imbalances between capacity and demand. The pressing, increasing demand for air transport registered in the last decade not only has already produced challenging delay management practices, but also fostered projections of even worse scenarios. EUROCONTROL [@statfor_growth_2018], for example, argued that delays in Europe could reach up to 20 minutes per flight in 2040, in stark contrast to the 12 minutes per flight, as registered in 2016.

@@

On March 11, 2020, the World Health Organisation has declared the novel coronavirus (COVID-19) outbreak a global pandemic.
The COVID-19 crisis posed a completely different, unexpected, and inverted challenge.
Communicable disease control resulted in massive restrictions on international and regional air traffic and passenger travel. 
Demand for air transport dropped as low as 90% of the previous "normal" in many places. Where the lack of capacity was previously the issue, now the lack of demand threatened the ATM system stability. 
In the financial perspective, airlines and airports had to deal with an unprecedented decrease in incomes. As a result, air navigation providers collected less fees for their services, due to significantly fewer flights. 
In the operational perspective, pilots and air traffic controllers practiced less. The problems and obstacles cascaded into many other dimensions.

@@

The unprecedented decline in air traffic demand resulted in severe financial strains on the air transport industry. 
The financial support or lack thereof for airlines and airports has been widely covered in the media. 
A variety of studies showed the interplay or consequences of the travel constraints. 
To date, lower attention was given to the inherent change in terms of air transport services.
Hence, the current scenario is a proper moment to further investigate the concept of resilience.

@@

* provide some descriptive text about COVID and government reactions (DONE BELOW!):

For example, in March 12, 2020, the United States established the suspension of most travels from the European Schengen area (BBC, 2020). A few days later, European Comission President Ursula von der Leyen announced a similar measure, proposing for EU Member States to apply restrictions on non-essential travel. By that time, several countries, such as Austria, Germany, and Poland had already implemented border check controls, while Portugal, Italy, and Spain have imposed lockdowns (Nielsen, 2020). At the same time, Australia imposed a severe arrival block on all non-citizens and non-residents at its airports (Worthington & Snape, 2020). In South America, Brazil, Chile, Argentina and other countries also followed the trend and closed their borders, totally or partially (France24, 2020). As a result, an unparalleled air traffic decline was seen worldwide (ICAO, 2021).

```{r, out.width='50%', fig.align='center'}
### It's not really open data, but shows the unprecedented decline mentioned above.###
#Use this only if you want...#

knitr::include_graphics('./figures/unprecedented_decline.jpg')

#Please add the references below to the bib:

#BBC. (2020, March 12). Coronavirus: Trump suspends travel from Europe to US. https://www.bbc.com/news/world-us-canada-51846923

#France24. (2020, March 16). Chile closes borders as Latam grounds flights over pandemic. https://www.france24.com/en/20200316-chile-closes-borders-as-latam-grounds-flights-over-pandemic

#International Civil Aviation Organization. (2021, July 13). Effects of novel Coronavirus (COVID‐19) on civil aviation:Economic impact analysis. https://www.icao.int/sustainability/Documents/COVID-19/ICAO_Coronavirus_Econ_Impact.pdf

#Nielsen, N. (2020, March 17). Commission proposes 'non-essential travel' EU entry ban. EU Observer. https://euobserver.com/coronavirus/147764

#Worthington, B., & Snape, J. (2020, March 19). Australia blocks arrival of all non-citizens, non-residents in expanded coronavirus travel ban. ABC News. https://www.abc.net.au/news/2020-03-19/coronavirus-non-resident-travel-ban-australia/12071640

```

* any more key milestones relevant for this study, e.g. US/EU stop, Lufthansa/BA/KLM/Air France stopping to operate to China as of late January

@@

Recently, the term is also used frequently on the political and strategic level. Policymakers, operational experts, and academics concur that the concept of resilience plays a major role when addressing and assessing the extent to which organisations and systems are prepared and capable to respond to and recover from disruptions.

@@

-   political level: The continuity and sustainability of a stable air traffic system supports and affects other dimensions of political interest, such as commercial relations, tourism industry, public health policies, among others. As a result, at a political level, it is of the utmost importance that the air navigation system remains functional. In that understanding, political acts were necessary in face of the pandemic crisis effects on aviation. For example, many governments injected financial support in the airline sector in order to help  them survive the worst periods. In addition, measures such as tax deferrals to air navigation services providers were also adopted. Therefore, understanding how the air navigation system can exist in a more resilient manner is certainly beneficial for many political agents.

-   operational level: While efficiency, delay management, and capacity constraints were previously the main problems affecting resilience, now the concept must be broadened to an inverted scenario. Airlines, ANSPs, airports, and other stakeholders in the industry must remain and manage resilience not only in order to recover from disruptions caused by excessive demand, but also from insufficient demand. If delays are no longer the main threat, others arise, such as insufficient funding, lack of training, and traffic unpredictability.

>>>>

---
title: "Assessing the Global COVID19 Impact on Air Transport with Open Data"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
# failed to render and I could not finally track what kills the pdf
 # bookdown::pdf_book:
 #   base_format: rticles::ieee_article
  bookdown::word_document2:
  #   reference_docx: tmpl_Conference-template-letter.docx
    default
bibliography: "DASC2021.bib"
affiliation:
  ## use one only of the following
  # author-columnar: true         ## one column per author
  institution-columnar: true  ## one column per institution  
  # wide: true          ## one column wide author/affiliation fields
  institution:
    - name: EUROCONTROL
      department: Performance Review Unit
      location: Brussels (Belgium)
      #mark: 1
      author:
        - name: Rainer Koelle
          email: rainer.koelle@eurocontrol.int
    - name: DECEA
      department: Subdepartment of Operations 
      location: Rio de Janeiro (Brazil)
      #mark: 2
      author:
        - name: Fabio Lourenco Carneiro Barbosa
          email: barbosaflcb@fab.mil.br
abstract: |
  This paper approaches the impact of the pandemic as a massive service disruption of the pre-pandemic global     connectivity and regional air transport networks. In particular, the project aims to provide data analytical evidence for policy success and transformation of the air transportation system. As an aspirational goal, the industry aims to recover in a “greener” manner. The project builds on openly available data sets. The paper will be produced in a reproducible manner making the data, code, and its processing available to interested reseachers and practitioners. The open assessment will provide policy makers with a tool to assess the reaction to local or regional measures.
  <!-- motivation -->
  <!-- short background - why important -->
  <!-- approach chosen / research method -->
  <!-- experimental work -->
  <!-- results obtained -->
  <!-- conclusions / key take-away --> 
---

@@

For many years, many concerns of the global air traffic management community has been directed to the evident problem of imbalances between capacity and demand. The pressing, increasing demand for air transport registered in the last decade not only has already produced challenging delay management practices, but also fostered projections of even worse scenarios. EUROCONTROL [@statfor_growth_2018], for example, argued that delays in Europe could reach up to 20 minutes per flight in 2040, in stark contrast to the 12 minutes per flight, as registered in 2016.

@@

On March 11, 2020, the World Health Organisation has declared the novel coronavirus (COVID-19) outbreak a global pandemic.
The COVID-19 crisis posed a completely different, unexpected, and inverted challenge.
Communicable disease control resulted in massive restrictions on international and regional air traffic and passenger travel. 
Demand for air transport dropped as low as 90% of the previous "normal" in many places. Where the lack of capacity was previously the issue, now the lack of demand threatened the ATM system stability. 
In the financial perspective, airlines and airports had to deal with an unprecedented decrease in incomes. As a result, air navigation providers collected less fees for their services, due to significantly fewer flights. 
In the operational perspective, pilots and air traffic controllers practiced less. The problems and obstacles developed into many other dimensions.

@@

The unprecedented decline in air traffic demand resulted in severe financial strains on the air transport industry. 
The financial support or lack thereof for airlines and airports has been widely covered in the media. 
A variety of studies showed the interplay or consequences of the travel constraints. 
To date, lower attention was given to the inherent change in terms of air transport services
Hence, the current scenario is a proper moment to further investigate the concept of resilience.

@@

* provide some descriptive text about COVID and government reactions
* any more key milestones relevant for this study, e.g. US/EU stop, Lufthansa/BA/KLM/Air France stopping to operate to China as of late January

@@

Recently, the term is also used frequently on the political and strategic level. Policymakers, operational experts, and academes concur that the concept of resilience plays a major role when addressing and assessing the extent to which organisations and systems are prepared and capable to respond to and recover from disruptions.

@@

-   political level: ADD SOME DESCRIPTION

-   operational level: ADD SOME DESCRIPTION
